---
title: 'Final Project - Step 3'
subtitle: "PSTAT126: Linear Regression" 

author:
  - name: "Ali Abuzaid"
    affiliations:
      - name: "Winter 2025"
affiliation-title: "Quarter"
format: 
 pdf: 
    latex_engine: xelatex
    geometry: "a4paper, margin=0.5in"
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
knitr::opts_chunk$set(echo =  FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
library(knitr)
library(dplyr)
library(ggplot2)
```


:::callout
<div style="text-align: center">
<span style="color: blue;"> **STUDENT NAME **</span> 
</div>
- Angelina Song    (angelinasong)
- Tasha Lee    (tashalee)
- Abbas Maheryar    (amaheryar)
- Maya Hoffstadt    (mayahoffstadt)
- Cyrus Mohammadian    (cmohammadian)
:::

# Abstract
The data displayed heteroscedasticity in both hypothesis 1 and hypothesis 2.  In both cases though, we reject the null hypotheses, asserting that there is a relationship between physical activity and BMI and finding that 3 different consumption habits have an effect on BMI as well. We may consider non-linear models or a larger data-set in the future such that we could perform analysis on a specific portion of the data, due to different ages and groups having different habits that could then overlap.

# Introduction
Body Mass Index (BMI) is commonly believed to be influenced by various lifestyle and dietary factors, including food consumption habits, hydration levels, alcohol intake, and frequency of physical activity. This study aims to analyze the relationship between different consumption and activity behaviors and BMI through linear regression to determine whether any particular habit has a significant impact on BMI. Using a dataset containing information on dietary patterns from the [UCI Machine Learning Repository](https://archive.ics.uci.edu/dataset/544/estimation+of+obesity+levels+based+on+eating+habits+and+physical+condition), we explore the link between consumption habits and BMI.

In this model, we include predictors such as the frequency of consuming high-caloric foods (FAVC), eating between meals (CAEC), alcohol consumption (CALC), and daily calorie monitoring (SCC), water intake (CH2O), vegetable consumption (FCVC), and frequency of physical exercise (FAF).

### Hypothesis
We would like to focus on two key research questions:
1) Does frequency of physical activity have any significant impact on BMI? (SLR)

$$
\begin{aligned}
H_0&: \text{Frequency of physical activity has no significant impact on BMI} \\
H_A&: \text{Frequency of physical activity has a significant impact on BMI}
\end{aligned}
$$
2) Do particular consumption habits have any significant impact on BMI? (MLR)
$$
\begin{aligned}
H_0&: \text{No particular consumption habit has any significant effect on BMI} \\
H_A&: \text{At least 1 consumption habit has a significant effect on BMI}
\end{aligned}
$$

# Data processing

For our dataset, we did not have any missing or null values. The first form of processing we did was to create a new column called `BMI` for the continuous variable of body-mass index.

Additionally, as per Professor Abuzaid's suggestion, we used functions `rnorm()`, `runif()`, and `rexp()`, to turn integer/categorical variables from the dataset into real continuous variables. We used the paper that came with the dataset to correctly map the right values for each category.

Finally, we combined the "Always" level in CALC into the "Frequent" level due to a lack of entries in "Always." We combined the "Automobile" and "Motorbike" levels in MTRANS into "Personal_Vehicle" due to a lack of entries in "Motorbike." Lastly, we combined the "Walking" and "Bike" levels in MTRANS into "Active_Transportation" due to a lack of entries in "Bike."

# Exploratory Data Analysis
### Null values
For our dataset, we did not have any missing or null values. 

### Boxplots & Scatterplots (MLR, Hypothesis 1)

For each variable in the multiple linear regression, we first created boxplots for categorical predictors and scatterplots for continuous predictors to display and to gain a better understanding the individual relationships between our variables.

```{r}
data <- read.csv('../data/data-cleaned.csv')
# categorical variables 
par(mfrow = c(2, 3), mar = c(2, 1, 2, 1), oma = c(1, 1, 1, 1))
boxplot(BMI ~ FAVC, data = data, col = "lightgreen", 
        main = "BMI vs FAVC", 
        xlab = 'Do you eat high caloric food frequently?')

boxplot(BMI ~ CAEC, data = data, col = "lightblue", 
        main = "BMI vs CAEC",
        xlab = 'Do you eat any food between meals?')


boxplot(BMI ~ CALC, data = data, col = "pink", 
        main = "BMI vs CALC",
        xlab = 'How often do you drink alcohol?')


boxplot(BMI ~ SCC, data = data, col = "burlywood", 
        main = "BMI vs SCC",
        xlab = 'Do you monitor the calories you eat daily?')

# continuous variables
model_ch2o <- lm(BMI ~ CH2O, data = data)
plot(BMI ~ CH2O, data = data, col = "cadetblue2", main = "BMI vs CH20",
        xlab = 'How often do you drink water?')
abline(model_ch2o, col = "blue", lwd = 2)

model_fcvc <- lm(BMI ~ FCVC, data = data)
plot(BMI ~ FCVC, data = data, col = "darkolivegreen1", main = "BMI vs FCVC",
        xlab = 'Do you usually eat vegetables in your meals?')
abline(model_fcvc, col = "darkgreen", lwd = 2)



```
### Boxplot for Simple Linear Regression, Hypothesis 1
```{r}
plot(BMI ~ FAF, data = data, col = "mediumpurple1", main = "BMI vs FAF",

        xlab = 'How often do you have physical activity??')

model_FAF <- lm(BMI ~ FAF, data = data)

abline(model_FAF, col = "mediumpurple4", lwd = 2)
```


From this EDA, we can see that for the categorical variables already have some effect on the BMI.

Predictors associated with higher BMI:

- frequently drinking alcohol (CALC)
- not counting calories (SCC)
- frequent consumption of high calorie food (FAVC)
- low frequency of physical activity (FAF)




### Correlation plot of BMI vs CH2O and FCVC
```{r}
#taking the correlation plot of the two continuous variables

library(corrplot)
corrplot_data <- data[, c("BMI", "CH2O", "FCVC")]
cor_matrix <- cor(corrplot_data)
corrplot(cor_matrix, method="number")
```
We also took a look at the correlation between the continuous predictors and the BMI. Here we see that these factors have a low value of multi-collinearity, which makes it suitable for the creation of our multiple linear regression model later.

# Explain your regression model

## Hypothesis 1
First, we can investigate whether frequency of physical activity has an impact on BMI. Our hypothesis are as follows:
$$
\begin{aligned}
H_0&: \beta_1=0\\
H_A&: \beta_1\neq0 \text{ ; where }\beta_1\text{ is the slope coefficient}
\end{aligned}
$$

Thus, the null hypothesis states that there is no linear relationship between BMI and frequency of physical activity and the alternative hypothesis is that there exists a linear relationship between BMI and frequency of physical activity.
Our linear model is as follows
$$ BMI=\beta_0+\beta_1FAF+e $$
```{r}
obesityData<- read.csv('../data/data-cleaned.csv')
# Fit linear model
BMIonFAF <- lm(BMI ~ FAF, data=obesityData)

# Get the summary of the model
BMIonFAF_summary <- summary(BMIonFAF)

# Format the coefficients and p-values
formatted_coefficients <- matrix(as.character(signif(
  BMIonFAF_summary$coefficients, 4)), ncol = 4)

colnames(formatted_coefficients) <- c("Estimate", "Std. Error",
                                      "t value", "p-value")
rownames(formatted_coefficients) <- c('Intercept',
                                      'Frequency of\nphysical activity')


kable(formatted_coefficients,
      caption = 'Fig. X: Simple Linear Regression of BMI on Frequency of Physical Activity')
```
The results of the simple linear model can be shown above in Figure X. At $\alpha=0.05$, we observe that our p-value, the probability that we observe a sample as or more extreme than our observed values given there is no relationship between BMI and frequency of physical activity is $p < 6.159e^{-36}$, or essentially 0. Thus, we reject our null hypothesis given we have statistically significant evidence that frequency of exercise impacts BMI. The slope coefficient $\hat{\beta_1}=-1.689$ states that for every one day per week increase in frequency of physical activity, our model predicts that BMI will decrease by 1.689 kilograms per meter.

Next, we can assess the goodness of fit of our model where $R^2=$ `r round(summary(BMIonFAF)$r.squared, 4)`. Thus, our model of BMI on frequency of physical activity accounts for 3.15% of total variability in BMI. As this value is rather low, we can examine why our model may not be a good fit using residual plots.
```{r}
# Generate residual plot
# Create residuals and fitted values
residuals <- resid(BMIonFAF)
fitted_values <- fitted(BMIonFAF)

data.frame(fitted_values, residuals) %>% 
  ggplot(aes(x = fitted_values,
             y = residuals)) +
  geom_point() +
  labs(title = "Fig. Y: Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()
```
Figure Y shows the spread of residuals. We should notice that there are four densely populated regions of residuals that may be due to rounding of answers. Also, notice how fitted values below 28 do not have residuals larger than 10 as opposed to other fitted values. As we are able to draw on some patterns in our residual plot, it may be beneficial to use a different model or fit our model to jittered values to eliminate the rounding effect in the future.

### Conclusion

We conclude at the significance level $\alpha < 0.05$ that there is a significant relationship between BMI and frequency of physical exercise, which is modeled by a nonzero coefficient in the linear model. So, we reject the null hypothesis. Moreover, while this model is saying that there is a significant relationship between these two variables that we cannot neglect, the simple linear regression model still leaves a lot of variance in the data unexplained, making us unable to accurately predict BMI with only the knowledge of one's physical activity habits.


## Hypothesis 2

### Initial Regression Analysis
```{r}
data <- read.csv("../data/data-cleaned.csv")
# lambda <- 0.6
# data$BMI <- (data$BMI^lambda - 1) / lambda
model_eating <- lm(BMI~Gender+Age+FAVC+ FCVC+ NCP+ CAEC+ CH2O+ SCC+ CALC,data)
s <- summary(model_eating)
knitr::kable(matrix(
  c(s$fstatistic[[1]],s$fstatistic[[2]],s$fstatistic[[3]],
    pf(s$fstatistic[[1]],s$fstatistic[[2]],s$fstatistic[[3]],
       lower.tail = FALSE),s$adj.r.squared),byrow = TRUE,nrow = 1
), col.names = c("F-stat","df1","df2","p-val", "R^2"),
caption="Overall Model Performance", digits = 200)
knitr::kable(s$coefficients, caption="Coefficient Analysis", digits=200)

```

### Model Adjustments
First, we drop insignificant parameters from the model and combine certain levels in categorical variables.
```{r}
data$CALC[data$CALC=="Sometimes"] <- "Yes"
data$CALC[data$CALC=="Frequently"] <- "Yes"

data$CAEC[data$CAEC=="Sometimes"] <- "Rarely"
data$CAEC[data$CAEC=="no"] <- "Rarely"

model_eating_u <- lm(BMI~Gender+Age+FAVC+ FCVC+ CAEC+ CH2O+ SCC+ CALC,data) # combined levels with 8 predictors
s<-summary(model_eating_u)
knitr::kable(matrix(
  c(s$fstatistic[[1]],s$fstatistic[[2]],s$fstatistic[[3]],
    pf(s$fstatistic[[1]],s$fstatistic[[2]],s$fstatistic[[3]],
       lower.tail = FALSE),s$adj.r.squared),byrow = TRUE,nrow = 1
), col.names = c("F-stat","df1","df2","p-val", "R^2"),
caption="Overall Model Performance", digits = 200)
knitr::kable(s$coefficients, caption="Coefficient Analysis", digits=200)
```

### Model Assumptions

Next, we check this updated model to see if the residuals match our assumptions of:
1. Normality

2. Homoscedasticity

3. Independence

4. Linearity

```{r, fig.width=8, fig.height=8}
par(mfrow = c(2, 2))
plot(model_eating_u)
```





Based on these plots, we can observe the following:
```{r echo=FALSE}
output <- matrix(
  c("Residuals vs. Fitted", "Funnel Shape indicates heteroscedasticity. No curvature to indicate nonlinearity",
    "Q-Q", "Vast majority of the points are on the line -- apears normal",
    "Scale-Location", "Funnel shape indicates heteroscedasticity",
    "residuals vs. Leverage", "Entry 345 possible influential point"),
  byrow=TRUE,nrow=4
)
knitr::kable(output,caption="Plot Interpretation")
```
Therefore, a transformation is in order. We begin by splitting our data to only include observations where the age is greater than 26.
```{r}
data <- data[data$Age>26,]
model_eating_u <- lm(BMI~Gender+Age+FAVC+ CAEC+ SCC+ CALC+ FCVC+ CH2O,data)
```
```{r, fig.width=8, fig.height=8}
par(mfrow = c(2, 2))
plot(model_eating_u)
```

We then attempt a box-cox transformation on the response variable.
```{r, fig.width=6, fig.height=4}
library(MASS)
suppressWarnings({
bc <- boxcox(model_eating_u, lambda=seq(-2,2,by=0.1),plotit=FALSE)
})
best_lambda <- bc$x[which.max(bc$y)]

data2 <- data
data2$BMI <- (data$BMI^best_lambda - 1) / best_lambda
model_eating_u <- lm(BMI~FAVC+ CAEC+ SCC+ CALC+ FCVC+ CH2O,data2)
cap <- paste("Residuals with lambda=", best_lambda)
plot(model_eating_u, which=c(1,2), caption = cap)
```
Next, we try Weighted Least Squared Regression:
```{r, fig.width=6, fig.height=4}
wt <- 1/lm(abs(model_eating_u$residuals)~model_eating_u$fitted.values)$fitted.values^2

wls_model <- lm(BMI~FAVC+ CAEC+ SCC+ CALC+ FCVC+ CH2O,data, weights = wt)
par(mfrow = c(2, 2))
plot(wls_model, caption="Weighted Regression Residuals")
```

# Results

Our results indicate that a multiple linear regression model regressing BMI on eating habits is overall statistically significant at the $\alpha=0.05$ significance level. Under the adjusted model summary statistics, the factors are all significant with p-value < 0.05. Therefore, we reject the null hypothesis and accept that alternate hypothesis that at least one predictor has a significant effect on BMI.

Notable results include that the coefficient for the binary predictor GenderMale is -0.6433362 with respect to the reference level (female). This means men generally have a BMI around half a point lower than women.

The continuous variable FCVC (Do you usually eat vegetables in your meals? ) resulted in coefficient of 3.580595. This indicates for each point increase in **not** eating vegetables is strongly associated with a BMI increase of 3.580595 points. 

The variable SCCyes (Monitoring the calories one eats daily) has a coefficient of -4.061563 compared to those who did not monitor calories, indicating monitoring calories was associated in ~ a 4-point decrease in BMI among study participants.

The variable CALCSometimes (How often do you drink alcohol? "sometimes") has a coefficient of 1.3937409, indicating those who drank alcohol sometimes had a BMI of 1.39 points *higher* than those who said "often," but those who did not drink alcohol at all (CALCno) had a BMI 0.908824 points *less* compared to those who said "often."

Overall, these findings suggest that dietary habits, calorie awareness, and alcohol consumption patterns play a significant role in BMI variations.

# Recommendations
For our multivariable regression model using eating habits, we had to go through multiple model transformations and adjustments, ending with a model that still did not meet certain residual assumptions. While filtering the data and only selecting observations that contained an Age of 26 or greater did remove a significant amount of heteroscedasticity, our residuals lost most of its normality in the process. Therefore, one recommendation of ours is to find a dataset that contains a sample more representative of the larger population of people. Another approach could be to only consider a specific population, such as adults below the age of 50 that do some form of physical activity consistently. 

With regards to our attempt at linear regression, an alternate approach could be to consider non-linear models, such as a neural network. Many of the residual assumptions we were struggling to meet would no longer be required.  So, this could potentially capture more variance in the data and decrease the Mean Standard Error. 

# Conclusion
In conclusion we can say that men generally have a lower BMI by half a point, likely due to physiology.  Eating vegetables and monitoring calories typically results in a lower BMI than otherwise.  It his hard to say how much of an effect that alcohol consumption has, but both "often" and "sometimes" had markedly higher, significant levels of BMI.  Several of our original variables and levels were insignificant, but we were able to combine them.  Even then, they do not appear to be a perfect fit, as we have a rather low $R^2$ even with the corrections in place.

The model itself, along with the data, likely wasn't able to accurately capture all of the factors that can be linked with BMI, and much of the original data was of integers and categorical, resulting in many factors and bloat.  In this situation, it would be best to try non-linear models and see if they can account for the data better, or a larger data-set that doesn't limit the majority of the answers to it to being integers such as "How often do you eat vegetables with meals?" getting the response "3".
