---
title: 'Final Project - Step 2 (20 Points)'
subtitle: "PSTAT126: Regression Analysis" 

author:
  - name: "Ali Abuzaid"
    affiliations:
      - name: "Winter 2025"
affiliation-title: "Quarter"
format: 
 pdf: 
    latex_engine: xelatex
    geometry: "letterpaper, margin=0.3in"
    code-fold: true
    code-line-numbers: true
    code-copy: true
    code-tools: true
    self-contained: true
    toc: false
    toc-location: left
    number-sections: true
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(message =  FALSE)
knitr::opts_chunk$set(warning =  FALSE)
knitr::opts_chunk$set(error =  FALSE)
knitr::opts_chunk$set(echo = FALSE)
bfcolor <- function(x, color) {
  if (knitr::is_latex_output()) {
    sprintf("\\textcolor{%s}{\\textbf{%s}}", color, x)
  } else if (knitr::is_html_output()) {
    sprintf("<span style='color: %s;'><b>%s</b></span>", color, x)
  } else x
}
```


:::callout
<div style="text-align: center">
<span style="color: blue;"> **STUDENT NAME **</span> 
</div>
- Tasha Lee    (tashalee)
- Abbas Maheryar    (amaheryar)
- Maya Hoffstadt   (mayahoffstadt)
- STUDENT 4    (NetID 4)
- STUDENT 5    (NetID 5)
:::


::: {.callout-caution}
## Due Date
The deadline for this step is **February 14, 2025**.
:::

:::{.callout-tip}
## Before Proceeding to Step 2
Please carefully review any comments or feedback on Step 1. If you have any questions or need clarification, reach out to the TA for assistance. Ensuring a clear understanding of Step 1 will help you complete the next phase successfully.
:::


# Step 2: Research Questions EDA, and Simple Linear Regression


## `r bfcolor("Research Questions", "red")` \ 

```{r echo=FALSE}
# Load packages
library(knitr)
library(ggplot2)
library(dplyr)

# Input data
obesityData <- read.csv("../data/BMI_data.csv")
```


**How does family history with obesity affect BMI?**


**How do eating habits (FAVC, FCVC, NCP, CAEC) affect BMI?**

**Question 3**

## `r bfcolor("Hypotheses", "red")` \ 

### Hypothesis 1
$$
\begin{aligned}
H_0&: \text{Family history does not affect BMI} \\
H_A&: \text{There is a significant relationship between family history and BMI}
\end{aligned}
$$


### Hypothesis 2
$$
\begin{aligned}
H_0&: \text{No particular eating habit has any effect on BMI} \\
H_A&: \text{At least 1 eating habit has a significant effect on BMI}
\end{aligned}
$$





## `r bfcolor(" Exploratory Data Analysis (EDA)", "red")` \  

### Data Cleaning
For our dataset, we did not have any missing or null values. The only form of processing we did was to create a new column called `BMI` for the continuous variable of body-mass index.



### Descriptive Statistics

Qualitative variable descriptive statistics:
(BMI, age, height, weight, FCVC, NCP, CH2O, FAF, TUE)
```{r echo=FALSE}
quant_vars <- obesityData %>%
  select(where(is.numeric))
summary_stats <-summary(quant_vars)

kable(summary_stats, col.names = colnames(quant_vars), 
      caption = "Summary Statistics for Quantitative Variables")
```


Quanitative variable descriptive statistics:
(Gender, family history, FAVC, CAEC, Smoke, SCC, CALC, MTRANS)



```{r echo=FALSE}

# Ensure that categorical variables are selected correctly and convert character to factors if needed
categorical_vars <- obesityData %>%
  mutate(across(where(is.character), as.factor)) %>%
  select(where(is.factor))


# Create an empty list to store summary tables for each variable
summary_list <- list()

# Loop through each categorical variable and calculate summary statistics
for (var in colnames(categorical_vars)) {
  
  # Create a frequency table for each variable
  freq_table <- table(categorical_vars[[var]])
  
  # Calculate the proportions (frequency)
  prop_table <- prop.table(freq_table)
  
  # Create a data frame to store counts and proportions
  summary_df <- data.frame(
    Category = names(freq_table),
    Count = as.integer(freq_table),
    Proportion = round(prop_table, 3)
  )
  
  # Add the summary table to the list
  summary_list[[var]] <- summary_df
}


# Loop through all categorical variables and print their summaries
for (var in colnames(categorical_vars)) {
  if (!is.null(summary_list[[var]]) && is.data.frame(summary_list[[var]])) {
    # Print the table for the variable using kable
    print(kable(summary_list[[var]][,c("Category", "Count", "Proportion.Freq")],
                caption = paste("Summary for", var)))
  } 
  }


```


### Data Visualization
```{r, fig.width=10, fig.height=5, echo=FALSE}
theme_update(text = element_text(size=10))
ggplot(obesityData, aes(x=MTRANS, y=BMI, 
                        fill = family_history_with_overweight)) +
  geom_boxplot() +
  ggtitle("Distributions of BMI based on family history of obesity") +
  labs(fill = "Fam Hist.")
```
```{r,fig.show="hold", fig.width=10,echo=FALSE}
library(patchwork)
theme_update(text = element_text(size=10))
a<-ggplot(obesityData, aes(x=factor(MTRANS))) +
  geom_bar() +
  xlab("Method of Transp.") +
  ggtitle("Method of Trans. Count Plot")
b<-ggplot(obesityData, aes(x=factor(family_history_with_overweight))) +
  geom_bar() +
  xlab("Family hist.") +
  ggtitle("Family Hist. of Obesity Count Plot")
a+b + plot_layout(ncol = 2) 
```
```{r, fig.width=10, fig.height=5, echo=FALSE}
ggplot(obesityData, aes(x=BMI)) +
  geom_histogram(aes(y=..density..),bins=30, fill="lightblue", color="black", alpha=0.6) +
  stat_function(fun=dnorm, args=list(mean=mean(obesityData$BMI),
                                     sd=sd(obesityData$BMI)),
                color="red", linewidth=1) +
  ggtitle("Histogram of BMI data with normal curve")

```
```{r}
continous <- obesityData[, c("Age", "NCP", "CH2O", 
                             "FAF", "TUE", "BMI")]
library(corrplot)
cor_mat <- cor(continous)
corrplot(cor_mat, method="number", "upper")
```






## `r bfcolor("Simple Linear Regression", "red")` \ 

### Hypothesis
We would like to perform simple linear regression of BMI (`BMI`) on frequency of physical activity (`FAF`). Our hypothesis are as follows:
$$
\begin{aligned}
H_0&: \beta_1=0\\
H_A&: \beta_1\neq0 \text{ ; where }\beta_1\text{ is the slope coefficient}
\end{aligned}
$$

Thus, the null hypothesis states that there is no linear relationship between BMI and frequency of physical activity and the alternative hypothesis is that there exists a linear relationship between BMI and frequency of physical activity.

### Assumptions of Linearity
Before proceeding with our simple linear regression, we must check the following assumptions:
1) Linearity
2) Independence
3) Homoscedasticity
4) Normality of residuals

### Test of hypothesis
Our linear model is as follows
$$ BMI=\beta_0+\beta_1Age+e $$
```{r}
# Fit linear model
BMIonFAF <- lm(BMI ~ FAF, data=obesityData)

# Get the summary of the model
BMIonFAF_summary <- summary(BMIonFAF)

# Format the coefficients and p-values
formatted_coefficients <- matrix(as.character(signif(
  BMIonFAF_summary$coefficients, 4)), ncol = 4)

colnames(formatted_coefficients) <- c("Estimate", "Std. Error",
                                      "t value", "p-value")
rownames(formatted_coefficients) <- c('Intercept',
                                      'Frequency of\nphysical activity')


kable(formatted_coefficients,
      caption = 'Fig. X: Simple Linear Regression of BMI on Fequency of Physical Activity')
```
The results of the simple linear model can be shown above in Figure X. At $\alpha=0.05$, we observe that our p-value, the probability that we observe a sample as or more extreme than our observed values given there is no relationship between BMI and frequency of physical activity is ____, or essentially 0. Thus, we reject our null hypothesis given we have statistically significant evidence that frequency of exercise impacts BMI. The slope coefficient $\hat{\beta_1}=-1.672$ states that for every one day per week increase in frequency of physical activity, our model predicts that BMI will decrease by 1.672 kilograms per meter.

We can also compute the confidence interval for $\hat{\beta_1}$ to further explore this relationship. The formula for the confidence interval is as follows
$$
\begin{aligned}
\hat{\beta_1}&\pm t_{\alpha/2}SE(\hat{\beta_1})\\
-1.672&\pm 1.96*0.2018\\
-1.672&\pm0.3955\\
\end{aligned}
$$
Thus, our confidence interval is [-2.0675, -1.2765].
```{r}
# Check calculation using R function
kable(confint(BMIonFAF),
      caption = 'Confidence Interval Calculation by R')
```

### Goodness-of-fit
Next, we can assess the goodness of fit of our model where $R^2=$ `r round(summary(BMIonFAF)$r.squared, 4)`. Thus, our model of BMI on frequency of physical activity accounts for 3.15% of total variability in BMI. As this value is rather low, we can examine why our model may not be a good fit using residual plots.
```{r}
# Generate residual plot
# Create residuals and fitted values
residuals <- resid(BMIonFAF)
fitted_values <- fitted(BMIonFAF)

data.frame(fitted_values, residuals) %>% 
  ggplot(aes(x = fitted_values,
             y = residuals)) +
  geom_point() +
  labs(title = "Fig. Y: Residuals vs Fitted Values",
       x = "Fitted Values",
       y = "Residuals") +
  theme_minimal()
```
Figure Y shows the spread of residuals. We should notice that there are four densely populated regions of residuals that may be due to rounding of answers. Also, notice how fitted values below 28 do not have residuals larger than 10 as opposed to other fitted values. As we are able to draw on some patterns in our residual plot, it may be beneficial to use a different model or fit our model to jittered values to eliminate the rounding effect in the future.

### Conclusion
